Direction-aware functional class scoring enrichment analysis of Infinium DNA methylation data

### 2024-01-18 manuscript/manuscript.Rmd
ms progress

We also wanted to know whether the LAM method could identify signatures that appear before patients are diagnosed
with a condition.
So we examined differential pathway methylation associated with incidence of 19 disease states.
Over all 19 conditions, there were 1570 significant pathways (FDR<0.05) (**Figure 9A**).
Liver cirrhosis has the most, with 464, followed by COPD with 400, while Alzheimer's has just one and COVID-19
hospitalisation had none.
A heatmap shows the enrichment scores for selected pathways across these incident conditions (**Figure 9B**).
Breast cancer incidence was associated with reduced methylation to the metal sequestration pathway.
Incident prostate cancer was associated with reduced methylation of the FGFR1 pathway.
Incident Parkinson's disease was associated with reduced methylation of aquaporins.
In COPD incidence, higher methylation of carbohydrate binding ficolins was observed together with reduced methylation
of the GDP mannose synthesis pathway.
Incident stroke was associated with higher methylation of the lipoxin synthesis pathway.
These results indicate that LAM can identify differentially methylated pathways in disease groups, and
detect altered methylation before a diagnosis has been made.

@@

@@

* Overall, we found that the FCS methods performed better across all simulation conditions.
Of the different FCS methods tested, LAM was superior in terms of precision and recall.
LAM also fits relatively easily into existing workflows, as most EWAS studies seem to use
limma in their analyses of differential probe methylation.
These analyses also consistently showed that non-parametric methods showed better recall with only a small decrease in precision, in accordance with some previous reports.
Summarisation to gene level only takes a few seconds with the efficient base R aggregation command,
and once aggregated, signatures can be analysed with existing packages like mitch.
Mitch does not require a large amount of system memory and due to its parallel architecture can
make use of multi-threaded processors, so it is not much slower than conducting ORA with existing methods.

* But it was interesting that ORA performed better when gene sets are smaller.
Given that KEGG and Reactome pathways consist of both large (>100) and small (<20) gene sets,
a hybrid FCS and ORA approach might increase sensitivity and recall.

* At 100 genes per set, LAM had superior sensitivity compared to GSA, which agrees with mitch's performance with simulated RNA-seq gene
expression data[@Kaspi2020-yx].
The improved sensitivity of this method provides a new opportunity for researchers to reanalyse previously conducted EWASs with contemporary pathway databases to better understand
the subtle signatures, as we have demonstrated with the chronological age and assisted reproduction examples.
The analyses conducted here used Reactome pathways, but this method is general purpose, so different gene sets can be used to examine various hypotheses.
Transcription factor target gene sets can be used to identify transcription factors associated with changes in pathway methylation.
MicroRNA target gene sets can identify potential associations between microRNA targets and DNA methylation patterns.
The MSigDB resource contains these and several other types of gene sets for exploration [@Liberzon2015-um].

* The differences in pathway results obtained with LAM and GSA from lung cancer data were striking, not only in the numbers of pathways identified, but also their relevance to the disease.
Interestingly the relationship between promoter methylation and gene expression was not always inverse as we had anticipated.
Joint enrichment analysis showed some genes with increasing promoter methylation together with increasing RNA expression (eg: "Negative regulation of AP2 transcription factors").
Likewise a few pathways showed decreases in methylation with decreases in RNA expression such as "Dectin family", "TLR regulation" and "glycan biosynthesis termination".
These results indicate that there is a complicated relationship between DNA methylation and gene expression in lung cancer and further work is required to appreciate the mechanisms involved.
* We have demonstrated that this method works with both HM450K and EPIC array data and can even facilitate the comparison of EWAS using different platforms.
We have provided an example workflow that researchers can use as a template.
It shows how to generate a probe-gene table from Bioconductor annotation datasets, how to update the defunct gene names and how to run the enrichment analysis.
Using this example, generating similar gene tables for the new EPIC 2.0 array and any future arrays will be straightforward, once their annotation sets are available through Bioconductor.
Updating defunct gene symbols is an important step as 2788 and 3253 gene symbols were updated on the HM450K and EPIC array annotations, which is approximately 10-14% of genes on these platforms.
We note that other packages for enrichment of Infinium array data to not by default address this, which could lead to a loss of these genes from pathway enrichment and a loss of sensitivity.
* The analysis of the large scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and incident disease.
The authors of the study used GOmeth for their pathway analysis and from the 33 models examined, significant findings were obtained for only two, prevalent diabetes and heart disease.
On contrast, LAM identified
Such work is important as it is known that methylation changes can be caused by disease states.
For example metabolic changes in diabetes are probably the cause of the observed methylation signatures.
It is perhaps unexpected that more significant pathways be identified from the incident disease arm of the study,

>>>>

We also wanted to know whether the LAM method could identify signatures that appear before patients are diagnosed
with a condition.
So we examined differential pathway methylation associated with incidence of 19 disease states.
Over all 19 conditions, there were 1570 significant pathways (FDR<0.05) (**Figure 9A**).
Liver cirrhosis has the most, with 464, followed by COPD with 400, while Alzheimer's has just one and COVID-19
hospitalisation had none.
A heatmap shows the enrichment scores for selected pathways across these incident conditions (**Figure 9B**).
Breast cancer incidence was associated with reduced methylation to the metal sequestration pathway.
Incident prostate cancer was associated with reduced methylation of the FGFR1 pathway.
Incident Parkinson's disease was associated with reduced methylation of aquaporins.
In COPD incidence, higher methylation of carbohydrate binding ficolins was observed together with reduced methylation
of the GDP mannose synthesis pathway.
Incident stroke was associated with higher methylation of the lipoxin synthesis pathway.

@@

To show that these findings are not the result of false positives from the LAM method, we randomised the
incidence profiles for all 19 conditions prior to enrichment analysis.
This was repeated 100 times, and these analyses of randimised data yielded few significant pathways.
Specifically, from 100 runs 64 yielded no significant pathways across 19 conditions.
Twenty eight repeats had fewer than five false positives and eight runs had more than five false positives.
Across the 100 repeats, the mean number of false positives was 1.95 and median was 0.
These results indicate that LAM can identify differentially methylated pathways in disease groups, and
detect altered methylation before a diagnosis has been made, with few false positives.

@@

* Overall, we found that the FCS methods performed better across all simulation conditions.
Of the different FCS methods tested, LAM was superior in terms of precision and recall.
LAM also fits relatively easily into existing workflows, as most EWAS studies seem to use
limma in their analyses of differential probe methylation.
The ALM method had only slightly worse performance as compared to LAM, but ALM requires users to aggregate the probe methylation values
for each sample for each gene prior to limma.
This extra aggregation step is computationally intensive and not a standard part of existing methylation analysis workflows.

Our analyses also consistently showed that non-parametric methods showed better recall with only a small decrease in precision, in accordance with some previous reports.
Summarisation of probe level t-statistics to gene level only takes a few seconds with the efficient base R `aggregate` command,
and once aggregated, signatures can be analysed with existing packages like mitch.
Mitch does not require a large amount of system memory and due to its parallel architecture can
make use of multi-threaded processors, so it is not much slower than conducting ORA with existing methods that are single threaded.

* An interesting observation is that ORA performed better when gene sets are smaller.
Given that popular pathway databases like KEGG and Reactome pathways consist of both large (>100) and small (<20) gene sets,
a hybrid FCS and ORA approach might increase sensitivity and recall.

* At 100 genes per set, LAM had superior sensitivity compared to GSA, which agrees with a prevous report of mitch's performance with
simulated RNA-seq gene expression data[@Kaspi2020-yx].
The improved sensitivity of the LAM method provides a new opportunity for researchers to reanalyse previously conducted EWASs with contemporary pathway databases to better understand
the subtle signatures, as we have demonstrated with the chronological age, assisted reproduction examples and 19 disease EWAS analyses.
The analyses conducted here used Reactome pathways, but LAM is general purpose, so different gene sets can be used to examine various hypotheses.
Transcription factor target gene sets can be used to identify transcription factors associated with changes in pathway methylation.
MicroRNA target gene sets can identify potential associations between microRNA targets and DNA methylation patterns.
The MSigDB resource contains these and several other types of gene sets for exploration [@Liberzon2015-um].

* The differences in pathway results obtained with LAM and GSA from lung cancer data were striking, not only in the numbers of pathways identified, but also their relevance to the disease.
Interestingly the relationship between promoter methylation and gene expression was not always inverse as we had anticipated.
Joint enrichment analysis showed some genes with increasing promoter methylation together with increasing RNA expression (eg: "Negative regulation of AP2 transcription factors").
Likewise a few pathways showed decreases in methylation with decreases in RNA expression such as "Dectin family", "TLR regulation" and "glycan biosynthesis termination".
These results indicate that there is a complicated relationship between DNA methylation and gene expression in lung cancer and further work is required to appreciate the mechanisms involved and distinguish which CpGs are critical for influencing gene expression and which are not.

* We have demonstrated that this method works with both HM450K and EPIC array data and can even facilitate the comparison of EWAS using different platforms.
We have provided an example workflow that researchers can use as a template.
It shows how to generate a probe-gene table from Bioconductor annotation datasets, how to update the defunct gene names and how to run the enrichment analysis.
Using this example, generating similar gene tables for the new Infinium Mouse and EPIC 2.0 arrays and any future arrays will be straightforward, once their annotation sets are available through Bioconductor.

* Updating defunct gene symbols is an important step as 2788 and 3253 gene symbols were updated on the HM450K and EPIC array annotations, which is approximately 10-14% of genes on these platforms.
We note that other packages for enrichment of Infinium array data do not by default address this, which could lead to a loss of these genes from pathway enrichment and diminished sensitivity.

* The analysis of the large scale disease EWAS shows that LAM can be readily applied to identify differential pathway methylation associated with prevalent and incident disease.
The authors of the study used GOmeth for their pathway analysis and from the 33 models examined, significant findings were obtained for only two, prevalent diabetes and heart disease.
On contrast, LAM identified 2,469 pathway associations (74.8 per model).
To demonstrate these are not the results of false positives, permutated profiles yielded on average 0.10 significant findings per model,
supporting the idea that LAM is identifying real pathway signatures.

* Distinguishing prevalent and incident signatures is important as it is known that methylation changes can be caused by disease states.
For example metabolic changes in diabetes are probably the cause of the observed methylation signatures.
It is perhaps unexpected that more significant pathways be identified from the incident disease arm of the study,
